# Introduction to Bayesian data analysis

Recall Bayes' rule:

\begin{theorem}\label{thm:bayes2}
  \textbf{Bayes' Rule}. Let $B_{1}$, $B_{2}$, ..., $B_{n}$ be mutually exclusive and exhaustive and let $A$ be an event with 
  $\mathbb{P}(A)>0$. Then 

\begin{equation}
\mathbb{P}(B_{k}|A)=\frac{\mathbb{P}(B_{k})\mathbb{P}(A|B_{k})}{\sum_{i=1}^{n}\mathbb{P}(B_{i})\mathbb{P}(A|B_{i})},\quad k=1,2,\ldots,n.\label{eq-bayes-rule}
\end{equation}	
\end{theorem}

When A and B are observable events, 
we can state the rule as follows:

\begin{equation}
p(A\mid B) = \frac{p(B\mid A) p(A)}{p(B)}
\end{equation}

Note that $p(\cdot)$ is the probability of an event.

When looking at probability distributions, we will encounter the rule in the following form. 

\begin{equation}
f(\theta\mid \hbox{data}) = \frac{f(\hbox{data}\mid \theta) f(\theta)}{f(y)}
\end{equation}

Here, $f(\cdot)$ is a probability density, not the probability of a single event.
$f(y)$ is called a ``normalizing constant'', which makes the left-hand side a probability distribution. 

\begin{equation}
f(y)= \int f(x,\theta)\, d\theta = \int f(y\mid \theta) f(\theta)\, d\theta
\end{equation}

If $\theta$ is a discrete random variable taking one value from the set $\{\theta_1,\dots,\theta_n \}$, then 

\begin{equation}
f(y)= \sum_{i=1}^{n} f(y\mid \theta_i) P(\theta=\theta_i)
\end{equation}


Without the normalizing constant, we have the relationship:

\begin{equation}
f(\theta\mid \hbox{data}) \propto f(\hbox{data}\mid \theta) f(\theta)
\end{equation}

Note that the likelihood $L(\theta; \hbox{data})$ (our data is fixed) is proportional to $f(\hbox{data}\mid \theta)$, and that's why we can refer to $f(\hbox{data}\mid \theta)$ as the likelihood in the following Bayesian incantation:


\begin{equation}
\hbox{Posterior} \propto \hbox{Likelihood}\times \hbox{Prior}
\end{equation}

Our central goal is going to be to derive the posterior distribution and then summarize its properties (mean, median, 95\% credible interval, etc.). 

Usually, we don't need the normalizing constant to understand the properties of the posterior distribution. That's why Bayes' theorem is often stated in terms of the proportionality shown above. 

Incidentally, this is supposed to be the moment of great divide between frequentists and Bayesians: the latter assign a probability distribution to the parameter, the former treat the parameter as a point value.

Two examples will clarify how we can use Bayes' rule to obtain the posterior. Both examples involve so-called conjugate priors, which are defined as follows:

\begin{quote}
Given the likelihood $f(x\mid \theta)$, if the prior $f(\theta)$ results in a posterior $f(\theta\mid x)$ that has the same form as $f(\theta)$, then we call $f(\theta)$ a conjugate prior.
\end{quote}


## Example 1: Binomial Likelihood, Beta prior, Beta posterior 

This is a contrived example, just meant to  provide us with a smooth entry into Bayesian data analysis. Suppose that an individual with aphasia answered 46 out of 100 questions correctly in a particular sentence comprehension task. The research question is, what is the probability that their average response is greater than 0.5, i.e., above chance.

The likelihood function will tell us $P(\hbox{data}\mid \theta)$:

```{r}
dbinom(46, 100, 0.5)
```

Note that 

\begin{equation}
P(\hbox{data}\mid \theta) \propto \theta^{46} (1-\theta)^{54}
\end{equation}

So, to get the posterior, we just need to work out a prior distribution $f(\theta)$. 

\begin{equation}
f(\theta\mid \hbox{data}) \propto f(\hbox{data}\mid \theta) f(\theta)
\end{equation}

For the prior, we need a distribution that can represent our uncertainty about the probabiliy $\theta$ of success. The Beta distribution is commonly used as prior for proportions. We say that the Beta distribution is conjugate to the binomial density; i.e., the two densities have similar functional forms.

The pdf is


\begin{equation*}
f(x)=  \left\{   
\begin{array}{l l}
       \frac{1}{B(a,b)} x^{a - 1} (1-x)^{b-1}  & \quad \textrm{if } 0< x < 1\\
       0 & \quad \textrm{otherwise}\\
\end{array} \right.
\end{equation*}

\noindent
where

\begin{equation*}
B(a,b) = \int_0^1 x^{a-1}(1-x)^{b-1}\, dx
\end{equation*}


In R, we write $X\sim\mathsf{beta}(\mathtt{shape1}=\alpha,\,\mathtt{shape2}=\beta)$. The associated $\mathsf{R}$ function is $\mathtt{dbeta(x, shape1, shape2)}$. 

The mean and variance are

\begin{equation} 
E[X]=\frac{a}{a+b}\mbox{ and }Var(X)=\frac{ab}{\left(a+b\right)^{2}\left(a+b+1\right)}.
\end{equation}

The Beta distribution's parameters a and b can be interpreted as (our beliefs about) prior successes and failures, and are called \textbf{hyperparameters}. Once we choose values for a and b, we can plot the Beta pdf. Here, we show the Beta pdf for three sets of values of a,b:

```{r,betas,fig.cap="\\label{fig:betaeg}Examples of the beta distribution with different parameter values."}
plot(function(x) 
  dbeta(x,shape1=1,shape2=1), 0,1,
      main = "Beta density",
              ylab="density",xlab="X",ylim=c(0,3))

text(.5,1.1,"a=1,b=1")

plot(function(x) 
  dbeta(x,shape1=3,shape2=3),0,1,add=T)
text(.5,1.6,"a=3,b=3")


plot(function(x) 
  dbeta(x,shape1=6,shape2=6),0,1,add=T)
text(.5,2.6,"a=6,b=6")
```

As Figure \ref{fig:betaeg} shows, as the a,b values are increased, the spread decreases.

If we don't have much prior information, we could use a=b=1; this gives us a uniform prior; this is called an uninformative prior or non-informative prior (although having no prior knowledge is, strictly speaking, not uninformative). If we have a lot of prior knowledge and/or a strong belief that $\theta$ has a particular value, we can use a larger a,b to reflect our greater certainty about the parameter. Notice that the larger our parameters a and b, the narrower the spread of the distribution; this makes sense because a larger sample size (a greater number of successes a, and a greater number of failures b) will lead to more precise estimates.

The central point is that the Beta distribution can be used to define the prior distribution of $\theta$.

Just for the sake of illustration, let's take four different beta priors, each reflecting increasing certainty. 

\begin{enumerate}
\item 
Beta(a=2,b=2)
\item
Beta(a=3,b=3)
\item 
Beta(a=6,b=6)
\item
Beta(a=21,b=21)
\end{enumerate}

Each reflects a belief that $\theta=0.5$, with varying degrees of (un)certainty. Now we just need to plug in the likelihood and the prior:

\begin{equation}
f(\theta\mid \hbox{data}) \propto f(\hbox{data}\mid \theta) f(\theta)
\end{equation}

The four corresponding posterior distributios would be:

\begin{equation}
f(\theta\mid \hbox{data}) \propto [\theta^{46} (1-\theta)^{54}] [\theta^{2-1}(1-\theta)^{2-1}] = \theta^{47} (1-\theta)^{55}
\end{equation}

\begin{equation}
f(\theta\mid \hbox{data}) \propto [p^{46} (1-p)^{54}] [p^{3-1}(1-p)^{3-1}] = p^{48} (1-p)^{56}
\end{equation}

\begin{equation}
f(\theta\mid \hbox{data}) \propto [\theta^{46} (1-\theta)^{54}] [\theta^{6-1}(1-\theta)^{6-1}] = \theta^{51} (1-\theta)^{59}
\end{equation}

\begin{equation}
f(\theta\mid \hbox{data}) \propto [\theta^{46} (1-\theta)^{54}] [\theta^{21-1}(1-\theta)^{21-1}] = \theta^{66} (1-\theta)^{74}
\end{equation}

We can now visualize each of these triplets of priors, likelihoods and posteriors. Note that I use the beta to model the likelihood because this allows me to visualize all three (prior, lik., posterior) in the same plot. The likelihood function is actually as shown in Figure \ref{fig:binomplot}:

```{r,binomlik,fig.cap="\\label{fig:binomplot}Binomial likelihood function."}
theta=seq(0,1,by=0.01)

plot(theta,dbinom(x=46,size=100,prob=theta),
     type="l",main="Likelihood")
```

We can represent the likelihood in terms of the Beta as well, as shown in Figure \ref{fig:betaforbinom}:

```{r,binomasbeta,fig.cap="\\label{fig:betaforbinom}Using the beta distribution to represent a binomial."}
plot(function(x) 
  dbeta(x,shape1=46,shape2=54),0,1,
              ylab="",xlab="X")
```


## Example 2: Poisson Likelihood, Gamma prior, Gamma posterior


This is also a contrived example. Suppose we are modeling the number of times that a 
speaker says the word ``the'' per day.

The number of times $x$ that the word is uttered in one day can be modeled by a Poisson distribution:

\begin{equation}
f(x\mid \theta) = \frac{\exp(-\theta) \theta^x}{x!}
\end{equation}

where the rate $\theta$ is unknown, and the numbers of utterances of the target word on each day are independent given $\theta$.

We are told that the prior mean of $\theta$ is 100 and prior variance for $\theta$  is 225. This information could be based on the results of previous studies on the topic.

In order to visualize the prior, we first fit a Gamma density prior for $\theta$ based on the above information. 

Note that we know that for a Gamma density with parameters a, b, the mean is  $\frac{a}{b}$ and the variance is
$\frac{a}{b^2}$.
Since we are given values for the mean and variance, we can solve for a,b, which gives us the Gamma density. 

If $\frac{a}{b}=100$ and $\frac{a}{b^2}=225$, it follows that
$a=100\times b=225\times b^2$ or $100=225\times b$, i.e., 
$b=\frac{100}{225}$.

This means that $a=\frac{100\times100}{225}=\frac{10000}{225}$.
Therefore, the Gamma distribution for the prior is as shown below (also see Figure \ref{fig1}):

\begin{equation}
\theta \sim Gamma(\frac{10000}{225},\frac{100}{225})
\end{equation}

```{r,fig.cap="\\label{fig1}The Gamma prior for the parameter theta."}
x<-0:200
plot(x,dgamma(x,10000/225,100/225),type="l",lty=1,main="Gamma prior",ylab="density",cex.lab=2,cex.main=2,cex.axis=2)
```


A distribution for a prior is \textbf{conjugate} if, multiplied by the likelihood, it yields a posterior that has the distribution of the same family as the prior. 

It turns out that the Gamma distribution is a conjugate prior for the Poisson distribution.
For the Gamma distribution to be a conjugate prior for the Poisson, the posterior needs to have the general form of a Gamma distribution. We derive this conjugacy result below. The proof just involves very simple algebra.

Given that 

\begin{equation}
\hbox{Posterior} \propto \hbox{Prior}~\hbox{Likelihood}
\end{equation}

and given that the likelihood is:

\begin{equation}
\begin{split}
L(\mathbf{x}\mid \theta) =& \prod_{i=1}^n \frac{\exp(-\theta) \theta^{x_i}}{x_i!}\\
          =& \frac{\exp(-n\theta) \theta^{\sum_i^{n} x_i}}{\prod_{i=1}^n x_i!}\\
\end{split}          
\end{equation}


we can compute the posterior as follows:

\begin{equation}
\hbox{Posterior} = \left[\frac{\exp(-n\theta) \theta^{\sum_i^{n} x_i}}{\prod_{i=1}^n x_i!} \right]
\left[\frac{b^a \theta^{a-1}\exp(-b\theta)}{\Gamma(a)}\right]
\end{equation}

We can disregard the terms $x!,\Gamma(a), b^a$,  which do not involve $\theta$, we have

\begin{equation}
\begin{split}
\hbox{Posterior} \propto &  \exp(-n\theta)  \theta^{\sum_i^{n} x_i} \theta^{a-1}\exp(-b\theta)\\
=& \theta^{a-1+\sum_i^{n} x_i} \exp(-\theta (b+n))
\end{split}
\end{equation}

We can now figure out the parameters of the posterior distribution, and show that it will be a Gamma distribution.

First, note that the Gamma distribution in general is $Gamma(a,b) \propto \theta^{a-1} \exp(-\theta b)$. So it's enough to state the above as a Gamma distribution with some parameters a*, b*.

If we equate $a^{*}-1=a-1+\sum_i^{n} x_i$ and $b^{*} = b+n$, we can rewrite the above as:

\begin{equation}
\theta^{a^{*}-1} \exp(-\theta b^{*})
\end{equation}

This means that $a^{*}=a+\sum_i^{n} x_i$ and $b^{*}=b+n$.
We can find a constant $k$ such that the above is a proper probability density function, i.e.:

\begin{equation}
\int_{-\infty}^{\infty} k \theta^{a^{*}-1} \exp(-\theta b^{*})=1
\end{equation}

Thus, the posterior has the form of  a Gamma distribution with parameters 
$a^{*}=a+\sum_i^{n} x_i, b^{*}=b+n$. Hence the Gamma distribution is a conjugate prior for the Poisson.

### Concrete example given data

Suppose we know that the number of ``the'' utterances is $115, 97, 79, 131$. We can derive the posterior distribution as follows.

The prior is Gamma(a=10000/225,b=100/225). The data are as given; this means that $\sum_i^{n} x_i = 422$ and sample size $n=4$.
It follows that the posterior is 

\begin{equation}
\begin{split}
Gamma(a^{*}= a+\sum_i^{n} x_i, b^{*}=b+n) =& 
Gamma(10000/225+422,4+100/225)\\
=& Gamma(466.44,4.44)\\
\end{split}
\end{equation}

The mean and variance of this distribution can be computed using the fact that the mean is $\frac{a*}{b*}=466.44/4.44=104.95$ and the variance is $\frac{a*}{b*^{2}}=466.44/4.44^2=23.66$.


```{r}
### load data:
data<-c(115,97,79,131)

a.star<-function(a,data){
  return(a+sum(data))
}

b.star<-function(b,n){
  return(b+n)
}

new.a<-a.star(10000/225,data)
new.b<-b.star(100/225,length(data))

### post. mean
(post.mean<-new.a/new.b) 
### post. var:
(post.var<-new.a/(new.b^2)) 
```

Now suppose you get one new data point:

```{r}
new.data<-c(200)
```

We can compute the parameters of the posterior Gamma distributions using the function we wrote above:

```{r}
new.a.2<-a.star(new.a,new.data)
new.b.2<-b.star(new.b,length(new.data))

### new mean
(new.post.mean<-new.a.2/new.b.2)
### new var:
(new.post.var<-new.a.2/(new.b.2^2))
```

Thus, new data can be used with the previous posterior distribution as prior, to compute the new posterior.

### The posterior is a weighted mean of prior and likelihood

We can express the posterior mean as a weighted sum of the prior mean and the maximum likelihood estimate of $\theta$.

The posterior mean is:

\begin{equation}
\frac{a*}{b*}=\frac{a + \sum x_i }{n+b}
\end{equation}

This can be rewritten as

\begin{equation}
\frac{a*}{b*}=\frac{a + n \bar{x}}{n+b}
\end{equation}

Dividing both the numerator and denominator by b:

\begin{equation}
\frac{a*}{b*}=\frac{(a + n \bar{x})/b }{(n+b)/b} = \frac{a/b + n\bar{x}/b}{1+n/b}
\end{equation}

Since $a/b$ is the mean $m$ of the prior, we can rewrite this as:

\begin{equation}
\frac{a/b + n\bar{x}/b}{1+n/b}= \frac{m + \frac{n}{b}\bar{x}}{1+
\frac{n}{b}}
\end{equation}

We can rewrite this as:

\begin{equation}
\frac{m + \frac{n}{b}\bar{x}}
{1+\frac{n}{b}} = \frac{m\times 1}{1+\frac{n}{b}} + \frac{\frac{n}{b}\bar{x}}{1+\frac{n}{b}}
\end{equation}

This is a weighted average: setting $w_1=1$ and 
$w_2=\frac{n}{b}$, we can write the above as:

\begin{equation}
m \frac{w_1}{w_1+w_2} + \bar{x} \frac{w_2}{w_1+w_2}
\end{equation}

A $n$ approaches infinity, the weight on the prior mean $m$ will tend towards 0, making the posterior mean approach the maximum likelihood estimate of the sample.

In general, in a Bayesian analysis, as sample size increases, the likelihood will dominate in determining the posterior mean.

Regarding variance, since the variance of the posterior is:

\begin{equation}
\frac{a*}{b*^2}=\frac{(a + n \bar{x})}{(n+b)^2} 
\end{equation}

as $n$ approaches infinity, the posterior variance will approach zero: more data will reduce variance (uncertainty). 

## Summary
 
 We saw two examples where we can do the computations to derive the posterior using simple algebra. There are several other such simple cases. However, in realistic data analysis settings, we cannot specify the posterior distribution as a particular density. We can only specify the priors and the likelihood. 
 
 For such cases, we need to use MCMC sampling techniques so that we can sample from the posterior distributions of the parameters.
 
We will discuss three approaches for sampling:

  - Gibbs sampling using inversion sampling
  - Metropolis-Hasting
  - Hamiltonian Monte Carlo
 
## MCMC sampling
 
### The inversion method for sampling
 
 This method works when we know the closed form of the pdf we want to simulate from and can derive the inverse of that function.

Steps:


\begin{enumerate}
\item Sample one number $u$ from $Unif(0,1)$. Let $u=F(z)=\int_L^z f(x)\, dx $ (here, $L$ is the lower bound of the pdf f).
\item Then $z=F^{-1}(u)$ is a draw from $f(x)$.
\end{enumerate}

#### Example 1: Samples from Standard Normal

Take a sample from the Uniform(0,1):

```{r}
u<-runif(1,min=0,max=1)
```

Let f(x) be a Normal density---we want to sample from this density. The inverse of the CDF in R is qnorm. It takes as input a probability and returns a quantile. 

```{r}
qnorm(u)
```

If we do this repeatedly, we will get samples from the Normal distribution (here, the standard normal).

```{r}
nsim<-10000
samples<-rep(NA,nsim)
for(i in 1:nsim){
  u <- runif(1,min=0,max=1)
  samples[i]<-qnorm(u)
}
hist(samples,freq=FALSE,main="Standard Normal")
```

#### Example 2: Samples from Exponential or Gamma

Now try this with the exponential with rate 1:

```{r}
nsim<-10000
samples<-rep(NA,nsim)
for(i in 1:nsim){
  u <- runif(1,min=0,max=1)
  samples[i]<-qexp(u)
}
hist(samples,freq=FALSE,main="Exponential")
```

Or the Gamma with rate and shape 1:

```{r}
nsim<-10000
samples<-rep(NA,nsim)
for(i in 1:nsim){
  u <- runif(1,min=0,max=1)
  samples[i]<-qgamma(u,rate=1,shape=1)
}
hist(samples,freq=FALSE,main="Gamma")
```


#### Example 3

Let $f(x) = \frac{1}{40} (2x + 3)$, with $0<x<5$. Now, we can't just use the family of q functions in R, because this density is not defined in R.

We have to draw a number from the uniform distribution and then solve for z, which amounts to finding the inverse function:

\begin{equation}
u = \int_0^z \frac{1}{40} (2x + 3)
\end{equation}

```{r}
u<-runif(1000,min=0,max=1) 

z<-(1/2) * (-3 + sqrt(160*u +9))
```

This method can't be used if we can't find the inverse, and it can't be used with multivariate distributions.

### Gibbs sampling

Gibbs sampling is a very commonly used method in Bayesian statistics. Here is how it works.

Let $\Theta$ be a vector of parameter values, let length of $\Theta$ be $k$. Let $j$ index the $j$-th iteration.

Algorithm:

\begin{enumerate}
\item Assign some starting values to $\Theta$:

$\Theta^{j=0} \leftarrow S$

\item 
Set $j \leftarrow j + 1$
\item 
\begin{enumerate}
\item[1.] Sample $\theta_1^j \mid \theta_2^{j-1}\dots \theta_k^{j-1}$.
\item[2.] Sample $\theta_2^j \mid \theta_1^{j}\theta_3^{j-1}\dots \theta_k^{j-1}$.

\vdots

\item[k.] Sample $\theta_k^{j} \mid \theta_1^{j}\dots \theta_{k-1}^{j}$.
\end{enumerate}
\item
Return to step 1.
\end{enumerate}

#### Example: A simple bivariate distribution

Assume that our bivariate (joint) density is:

\begin{equation}
f(x,y)= \frac{1}{28}(2x + 3y + 2)
\end{equation}

Using the methods discussed in the Foundations chapter, it is possible to analytically work out the conditional distributions from the joint distribution:

\begin{equation}
f(x\mid y)=  \frac{f(x,y)}{f(y)}= \frac{(2x + 3y + 2)}{6y+8}
\end{equation}

\begin{equation}
f(y\mid x)=  \frac{f(x,y)}{f(x)}= \frac{(2x + 3y + 2)}{4y+10}
\end{equation}

The Gibbs sampler algorithm is: 

\begin{enumerate}
\item
Set starting values for the two parameters $x=-5, y=-5$. Set j=0.
\item
Sample $x^{j+1}$ from $f(x\mid y)$ using inversion sampling. You need to work out the inverse of $f(x\mid y)$ and $f(y\mid x)$ first.
To do this, for $f(x\mid u)$, we have 
find $z_1$:

\begin{equation}
u = \int_0^{z_1} \frac{(2x + 3y + 2)}{6y+8}\, dx
\end{equation}

And for $f(y\mid x)$, we have to find $z_2$:

\begin{equation}
u = \int_0^{z_2} \frac{(2x + 3y + 2)}{4y+10} \, dy
\end{equation}

It doesn't matter if you can't solve this integral; the solution is given in the code below.
\end{enumerate}

```{r}
#R program for Gibbs sampling using inversion method 
## program by Scott Lynch, modified by SV:
x<-rep(NA,2000)
y<-rep(NA,2000) 
x[1]<- -5
y[1]<- -5

for(i in 2:2000)
{ #sample from x | y 
  u<-runif(1,min=0, max=1) 
  x[i]<-sqrt(u*(6*y[i-1]+8)+(1.5*y[i-1]+1)*(1.5*y[i-1]+1))-
    (1.5*y[i-1]+1) 
  #sample from y | x
u<-runif(1,min=0,max=1) 
y[i]<-sqrt((2*u*(4*x[i]+10))/3 +((2*x[i]+2)/3)*((2*x[i]+2)/3))- 
    ((2*x[i]+2)/3)
}
```

You can run this code to visualize the simulated posterior distribution. See Figure \ref{fig:posteriorbivariateexample}.

```{r fig.cap="\\label{fig:posteriorbivariateexample}Example of posterior distribution of a bivariate distribution."}
library(MASS)
bivar.kde<-kde2d(x,y)
persp(bivar.kde,phi=10,theta=90,shade=0,border=NA,
      main="Simulated bivariate density using Gibbs sampling")
```


A central insight here is that knowledge of the conditional distributions is enough to simulate from the joint distribution, provided such a joint distribution exists. 

## Hamiltonian Monte Carlo {#hmc}

Instead of Gibbs sampling, Stan uses this more efficient sampling approach. HMC works well for the high-dimensional models we will fit (hierarchical models).
Gibbs sampling faces difficulties with some of the complex hierarchical models we will be fitting later. HMC will always succeed for these complex models.

One limitation of HMC (which Gibbs sampling does not have) is that HMC only works with continuous parameters (not discrete parameters).

For our purposes, it is enough to know what sampling using MCMC is, and that HMC gives us posterior samples efficiently.
A good reference explaining HMC is @neal2011mcmc. However, this paper is technically very demanding. More intuitively accessible introductions are available via Michael Betancourt's home page: https://betanalpha.github.io/. In particular, this video is helpful:
https://youtu.be/jUSZboSq1zg.

### HMC demonstration 

The HMC algorithm takes as input the log density and the gradient of the log density. In Stan, these will be computed internally; the user doesn't need to do any computations.

For example, suppose the log density is $f(\theta) = - \frac{\theta^2}{2}$.
Its gradient is $f'(\theta) = -\theta$. Setting this gradient to 0, and solving for $\theta$, we know that the maximum is at 0. We know it's a maximum because the second derivative, $f''(\theta) = -1$, is negative. See Figure \ref{fig:logdensityexample}.  

This is the machinery we learnt in the foundations chapter (recall how we found MLEs in particular).

```{r fig.cap="\\label{fig:logdensityexample}Example log density."}
theta<-seq(-4,4,by=0.001)
plot(theta,-theta^2/2,type="l",main="Log density")
```

We first write down the Radford Neal algorithm for HMC; details can be ignored in the next piece of code below, which I got from [Jarad Niemi](https://github.com/jarad/)'s github repository. 

```{r}
## Radford Neal algorithm:
HMC_neal <- function(U, grad_U, e, L, current_theta) {
theta = current_theta
omega = rnorm(length(theta),0,1)
current_omega = omega
omega = omega - e * grad_U(theta) / 2
for (i in 1:L) {
theta = theta + e * omega
if (i!=L) omega = omega - e * grad_U(theta)
}
omega = omega - e * grad_U(theta) / 2
omega = -omega
current_U = U(current_theta)
current_K = sum(current_omega^2)/2
proposed_U = U(theta)
proposed_K = sum(omega^2)/2
if (runif(1) < exp(current_U-proposed_U+current_K-proposed_K))
{
return(theta)
}
else {
return(current_theta)
}
}

HMC <- function(n_reps, log_density, grad_log_density, tuning, initial) {
theta = rep(0, n_reps)
theta[1] = initial$theta
for (i in 2:n_reps) theta[i] = HMC_neal(U = function(x) -log_density(x),
grad_U = function(x) -grad_log_density(x),
e = tuning$e,
L = tuning$L,
theta[i-1])
theta
}
```

Then, we use the HMC function above to take 2000 samples from the posterior. 
We drop the first few (typically, the first half) samples, which are called warm-ups. The reason we drop them is that the initial samples may not yet be sampling from the posterior.


```{r}
theta <- HMC(n_reps=2000, 
             log_density=function(x) -x^2/2, 
             grad_log_density=function(x) -x, 
             tuning=list(e=1,L=1),
             initial=list(theta=0))
```

Figure \ref{fig:hmcsamples0} shows a **trace plot**, the trajectory of the samples over 2000 iterations. This is called a **chain**. When the sampler is correctly sampling from the posterior, we see a characteristic ``fat hairy caterpillar'' shape, and we say that the sampler has **converged**. You will see later what a failed convergence looks like.


```{r fig.cap="\\label{fig:hmcsamples0}An example of a trace plot."}
plot(theta,type="l",main="Trace plot of posterior samples")
```

When we fit Bayesian models, we will always run four parallel chains. This is to make sure that even if we start with four different initial values chosen randomly, the chains all end up sampling from the same distribution. When this happens, we see that the chains overlap visually, and we say that the chains are **mixing**.

Figure \ref{fig:hmcsamples} shows the posterior distribution of $\theta$. We are not discarding samples here because the sampler converges quickly in this simple example.


```{r fig.cap="\\label{fig:hmcsamples}Sampling from the posterior using HMC. The red curve shows the distribution Normal(0,1)."}
hist(theta, freq=F, 100,
     
     main="Posterior distribution of the parameter.",
     xlab=expression(theta))
curve(dnorm, add=TRUE, col='red', lwd=2)
```

In the modeling we do in the following pages, the Stan software will do the sampling for us. 

\newpage 

## Further readings

Some good books introducing Bayesian data analysis methods are the following:

  - McElreath, R. (2015). Statistical Rethinking. Texts in Statistical Science.
This book is currently the best textbook in existence, and assumes no calculus or linear algebra knowledge.

  - Kruschke, J. (2014). Doing Bayesian data analysis: A tutorial with R, JAGS, and Stan. Academic Press. This book is specifically designed for a psychology audience. I have only read parts of it and I find it very useful as a reference.

  - Lynch, S. M. (2007). Introduction to applied Bayesian statistics and estimation for social scientists. Springer Science & Business Media. This book assumes knowledge of calculus and linear algebra. It gives a classical, statistician's introduction to Bayes. I highly recommend this book to people who know some (undergraduate math level) calculus and linear algebra.

\begin{qbox}{\subsection{Exercises}}

\begin{enumerate}
  \item 
    Suppose we are given that the pdf of $\theta$, which ranges from 0 to 1, is proportional to 
$\theta^2$. This means that there is some constant c (the constant of proportionality) such that $1=c \int_0^1 \theta^2\, d\theta$.

\begin{enumerate}
\item Find c.
\item Find the mean, median (hint: what is the median in terms of quantiles?) and variance of the above pdf.
\item Find the 95\% credible interval, i.e., the lower and upper values in $P(lower < \theta < upper)= 0.95$.
\end{enumerate}

\item Exercise on the Binomial distribution:
Toss a coin 10 times and compute, using \texttt{pbinom}, the probability of getting the total numbers of heads you got, assuming that the coin is fair.

Hint: given sample size $n$, your assumed probability of a heads $prob$, and the number of heads you got $x$, the \texttt{pbinom} function delivers $P(X \leq x)$, the probability of getting $x$ heads or less. In other words, pbinom is the \textbf{cumulative distribution function} (textbooks often call this simply \textbf{distribution function}).

Here is how you can compute $P(X\leq x)$:

\begin{verbatim}
pbinom(x,size=n,p=prob)
\end{verbatim}

```{r,include=FALSE}
nheads<-8
pbinom(nheads,size=10,p=0.5)-pbinom((nheads-1),size=10,p=0.5)
```

Note that you have to compute $P(X=x)$!

Based on what everyone finds, write down the number of cases we have of each possible outcome, along with the theoretical probability. 


\item
Given 
$X \sim f(\cdot)$, where $f(\cdot)$ is (a) $Unif(0,10)$, (b) 
$N(\mu=100,\sigma^2=20)$, (c) 
$Binom(p=.6,n=20)$. 
Find the probability of $P(X<3), P(X>11), P(X=6)$ 
for each distribution.

\item
Plot the priors, likelihoods, and posterior distributions in the four Beta-Binomial cases discussed in the notes.

```{r,binomexample1,echo=F,include=F}
###lik:
plot(function(x) 
  dbeta(x,shape1=46,shape2=54),0,1,
              ylab="",xlab="X",col="red")

### prior:
plot(function(x) 
  dbeta(x,shape1=2,shape2=2), 0,1,
      main = "Prior",
              ylab="density",xlab="X",add=T,lty=2)

### posterior
plot(function(x) 
  dbeta(x,shape1=48,shape2=56), 0,1,
      main = "Posterior",
              ylab="density",xlab="X",add=T)

legend(0.1,6,legend=c("post","lik","prior"),
       lty=c(1,1,2),col=c("black","red","black"))
```

\end{enumerate}
\end{qbox}



